---
title: "R Notebook"
output: html_notebook
---

list

data <- read.csv("results.csv")

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
```{r}
test = read.table("data/leukemia/molecular/rnaseq/genes/FT-SA158752R.genes.results")

test = read.table("data/leukemia/molecular/rnaseq/genes/FT-SA158752R.genes.results", header = TRUE)
```
```{r}
test = test %>%
  select(gene_symbol,TPM)
```
```{r}
test2 = transpose(test)

#Step 1: List all files in directory
files <- list.files(path = "data/leukemia/molecular/rnaseq/genes", pattern = "\\.genes\\.results$", full.names = TRUE)
# Step 2: Read each file, select gene name + TPM, and rename TPM to sample ID
read_tpm <- function(file) {
  sample_id <- tools::file_path_sans_ext(basename(file))  
  # Extract file name without extension
  df <- read_tsv(file, show_col_types = FALSE) %>%
    select(gene_id = gene_id, !!sample_id := TPM)
  return(df)
}
# Step 3: Load and merge all by gene
merged_tpm <- reduce(lapply(files, read_tpm), full_join, by = "gene_id")

 List all TSV files in the folder
> files <- list.files(pattern = "\\.tsv$")
> 
> # Read and trim each file
> all_data <- lapply(files, function(file) {
+   df <- read_tsv(file)
+ 
+   # Standardize column names
+   colnames(df) <- tolower(gsub(" ", "_", colnames(df)))
+ 
+   # Select gene symbol and tpm/trm columns
+   df_selected <- df %>%
+     select(matches("gene.*symbol|gene.*name"), matches("tpm|trm"))
+ 
+   df_selected$file <- file  # optional: add source file name
+   return(df_selected)
+ })
> 
> # Combine all into one dataframe
> combined_data <- bind_rows(all_data)
> 
> # Preview the result
> head(combined_data)
> head(merged_tpm)
 #Step 4: Filter out >100 NAs
filtered_merged <- merged_tpm %>%
+     filter(rowSums(is.na(across(-gene_id))) <= 100)
> head(filtered_merged)

library(ggplot2)
> library(tibble)

expr_matrix <- filtered_merged %>%
+     column_to_rownames("gene_id") %>%
+     t() %>%
+     as.data.frame()

# Transpose: genes as rows → samples as rows
> expr_matrix <- filtered_merged %>%
+     column_to_rownames("gene_id") %>%
+     t() %>%
+     as.data.frame()
> 
> # Remove genes (columns) with zero variance
> expr_matrix <- expr_matrix[, apply(expr_matrix, 2, function(x) sd(x, na.rm = TRUE) > 0)]
> 
> # Replace NAs with 0 in the entire expression matrix
> expr_matrix[is.na(expr_matrix)] <- 0
> 
> # Double-check if there are still NAs
> sum(is.na(expr_matrix))  # Should return 0 now
[1] 0
> 
> # Now run PCA
> pca_result <- prcomp(expr_matrix, scale. = TRUE)
> 
> # Create a data frame with PCA results for plotting
> pca_df <- as.data.frame(pca_result$x)
> pca_df$sample <- rownames(pca_df)  # Add sample names
> 
> # Optional: Cluster the samples using k-means (adjust number of clusters if needed)
> set.seed(42)
> k_clusters <- 3  # You can change this number if needed
> pca_df$cluster <- as.factor(kmeans(pca_df[, 1:2], centers = k_clusters)$cluster)
> 
> # Plot the PCA results
> ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster, label = sample)) +
+     geom_point(size = 3, alpha = 0.8) +
+     geom_text(vjust = 1.5, hjust = 1.1, size = 2.5) +
+     labs(title = "PCA of RNA-seq Samples", x = "PC1", y = "PC2") +
+     theme_minimal() +
+     theme(legend.title = element_blank())
> 
> ggplot(pca_df, aes(x = PC1, y = PC2)) +
+     geom_point(size = 3, alpha = 0.8) +
+     labs(title = "PCA of Patients by Gene Expression",
+          x = paste0("PC1 (", round(100 * summary(pca_result)$importance[2, 1], 1), "%)"),
+          y = paste0("PC2 (", round(100 * summary(pca_result)$importance[2, 2], 1), "%)")) +
  theme_minimal()
Create a data frame with PCA results
> pca_df <- as.data.frame(pca_result$x)
> pca_df$sample <- rownames(pca_df)
> 
> # Optional: K-means clustering
> set.seed(42)
> k_clusters <- 3
> pca_df$cluster <- as.factor(kmeans(pca_df[, 1:2], centers = k_clusters)$cluster)
> 
> # Calculate % variance explained for PC1 and PC2
> pc1_var <- round(100 * summary(pca_result)$importance[2, 1], 1)
> pc2_var <- round(100 * summary(pca_result)$importance[2, 2], 1)
> 
> # Plot with dynamic axis labels
> ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster, label = sample)) +
+     geom_point(size = 3, alpha = 0.8) +
+     geom_text(vjust = 1.5, hjust = 1.1, size = 2.5) +
+     labs(
+         title = "PCA of RNA-seq Samples",
+         x = paste0("PC1 (", pc1_var, "%)"),
+         y = paste0("PC2 (", pc2_var, "%)")
+     ) +
+     theme_minimal() +
+     theme(legend.title = element_blank())

#Adding Avatarid
rename_vector <- setNames(id_map$ORIENAvatarKey, id_map$RNASeq)
> gene_ids <- filtered_merged$gene_id
> expr_only <- filtered_merged[, -1]
> colnames(expr_only) <- rename_vector[colnames(expr_only)]
> filtered_avatar <- cbind(gene_id = gene_ids, expr_only)
> list(filtered_avatar)
ename_map <- setNames(id_map$ORIENAvatarKey, id_map$RNASeq)
> filtered_merged <- filtered_merged %>%
+     rename_with(~ recode(., !!!rename_map))
> gene_ids <- filtered_merged$gene_id
> expr_only <- filtered_merged[, -1]
> colnames(expr_only) <- rename_vector[colnames(expr_only)]
> filtered_avatar <- cbind(gene_id = gene_ids, expr_only)

```


```{r}
# Step 4: Filter out genes with >100 NAs across all samples
filtered_merged <- merged_tpm %>%
  filter(rowSums(is.na(across(-gene_id))) <= 100)
head(filtered_merged)
```


```{r}
library(ggplot2)
library(tibble)

# Step 1: Prepare the matrix for PCA (transpose so samples are rows)
expr_matrix <- filtered_merged %>%
  column_to_rownames("gene_id") %>%
  t() %>%
  as.data.frame()

# Step 2: Run PCA
pca_result <- prcomp(expr_matrix, scale. = TRUE)

# Step 3: Create a PCA dataframe for plotting
pca_df <- as.data.frame(pca_result$x)
pca_df$sample <- rownames(pca_df)  # Add sample names

# Optional: k-means clustering to color points by cluster
set.seed(42)
k_clusters <- 3  # Adjust number of clusters if needed
pca_df$cluster <- as.factor(kmeans(pca_df[, 1:2], centers = k_clusters)$cluster)

# Step 4: Plot PCA
ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster, label = sample)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text(vjust = 1.5, hjust = 1.1, size = 2.5) +
  labs(title = "PCA of RNA-seq Samples (filtered_merged)", x = "PC1", y = "PC2") +
  theme_minimal()
